---
title: "R Notebook"
#output: html_notebook

output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

# Downloading Data from the internet
Use `download.file()` that takes the url and the destination on your computer to save the file at. And note that

* If the file url starts with **http** you can use `download.file()`.
* It is also okay to use `download.file()` in case of **https** and **windows**.
* But on **Mac**, you might need to set `method="curl"` to download from **https** url.
* The download duration depends on the file size.

Example: download the 2006 microdata survey about housing for the state of Idaho.
```{r}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv"
download.file(url, "house_data.csv") ##save in the current dir, and name the file "data.csv"
```

List the files in the current directory
```{r}
list.files("./")
```

Use `date()` to get the downloading date.
```{r}
downloadDate <- date()
downloadDate
```

# Reading Excel Files
First download the file using `download.file()`
```{r}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FDATA.gov_NGAP.xlsx"
download.file(url, "data.xlsx",mode="wb")
date()
```

## Using `xlsx` Package

You can read the file using `read.xlsx` or `read.xlsx2`. You will need to download the `xlsx` package to use them.
```{r}
# Load the library
library(rJava)
library(xlsxjars)
library(xlsx)
readFile <- read.xlsx("data.xlsx", sheetIndex=1)
head(readFile)
```
 
You might face some difficulities when using the xlsx library. Here is the most famous error:
 
Error : .onLoad failed in loadNamespace() for 'rJava', details:
  call: inDL(x, as.logical(local), as.logical(now), ...)

  error: unable to load shared object 'C:/Users/me/Documents/R/win-library/2.13/rJava/libs/x64/rJava.dll':
  LoadLibrary failure:  %1 is not a valid Win32 application.

Error: package/namespace load failed for 'rJava'

**To solve that you will need to :**

* download the `rJava` package (if the error remains)
* check your R version `R.version()`, and download the corresponding java (32/64), from [here](https://www.java.com/en/download/win10.jsp).
* Finally, check if your Java is in `Program Files` or `Program Files (x86)`. Add the path to the **jvm.dll** to your PATH in **windows Environment**. If the java file is in `Program Files (x86)`, it means you have 32-bit version, and you can change the default version of your `Rstudio` from Tools >> Global options to 32 bit. For more information check [this](http://stackoverflow.com/questions/7019912/using-the-rjava-package-on-win7-64-bit-with-r).

### You can read specific rows and columns form the Excel file
```{r}
c <- 7:15
r <- 18:23
readSubset <- read.xlsx("data.xlsx", sheetIndex=1, colIndex = c, rowIndex = r)
readSubset
```

## Using `readxl` Package
Another way to read Excel files is to use `read_excel` from the `readxl` library. For more information check [this](http://stackoverflow.com/questions/7049272/importing-xlsx-file-into-r).

After installing the package..
```{r}
library(readxl)
dataFile <- read_excel("data.xlsx")
head(dataFile)
```

