---
title: "R Notebook"
#output: html_notebook

output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

# Downloading Data from the internet
Use `download.file()` that takes the url and the destination on your computer to save the file at. And note that

* If the file url starts with **http** you can use `download.file()`.
* It is also okay to use `download.file()` in case of **https** and **windows**.
* But on **Mac**, you might need to set `method="curl"` to download from **https** url.
* The download duration depends on the file size.

Example: download the 2006 microdata survey about housing for the state of Idaho.
```{r}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv"
download.file(url, "data.csv") ##save in the current dir, and name the file "data.csv"
```

List the files in the current directory
```{r}
list.files("./")
```

Use `date()` to get the downloading date.
```{r}
downloadDate <- date()
downloadDate
```

# Reading Excel Files
First download the file using `download.file()`
```{r}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FDATA.gov_NGAP.xlsx"
download.file(url, "data.xlsx",mode="wb")
date()
```

## Using `xlsx` Package

You can read the file using `read.xlsx` or `read.xlsx2`. You will need to download the `xlsx` package to use them.
```{r}
# Load the library
library(xlsx)
readFile <- read.xlsx("data.xlsx", sheetIndex=1)
```
Roses are <span style="color:red">red</span>, 
violets are <span style="color:blue">blue</span>.

